---
title: "Regession2"
author: "Manirath Wongsim"
date: ""
output: 
  html_document:
    fig_width: 7
    fig_height: 5
---

We'll begin by loading some packages.
```{r}
library(MASS)
library(plyr)
library(ggplot2)
```



```{r}
View(birthwt)

```





```{r}
colnames(birthwt)<-c("birthwt.below.2500","mother.age",
    "mother.weight","race","mother.smoke","privious.prem.labor","hypertension",
    "uterine.irr","physician.visits","birthwt.grams")

View(birthwt)
```

```{r}
library(plyr)
birthwt<-transform(birthwt,race=as.factor(mapvalues(race,c(1,2,3),c("white","black","other"))),
mother.smokes=as.factor(mapvalues(mother.smokes,c(0,1),c("no","yes"))),
hypertension=as.factor(mapvalues(hypertension,c(0,1),c("no","yes"))),
uterine.irr=as.factor(mapvalues(uterine.irr,c(0,1),c("no","yes"))),
birthwt.below.2500=as.factor(mapvalues(birthwt.below.2500,c(0,1),c("no","yes"))))
summary(birthwt)
```















Let's form our favourite birthwt data set.

```{r}
# Rename the columns to have more descriptive names
colnames(birthwt) <- c("birthwt.below.2500", "mother.age", "mother.weight", 
    "race", "mother.smokes", "previous.prem.labor", "hypertension", "uterine.irr", 
    "physician.visits", "birthwt.grams")

# Transform variables to factors with descriptive levels
birthwt <- transform(birthwt, 
            race = as.factor(mapvalues(race, c(1, 2, 3), 
                              c("white","black", "other"))),
            mother.smokes = as.factor(mapvalues(mother.smokes, 
                              c(0,1), c("no", "yes"))),
            hypertension = as.factor(mapvalues(hypertension, 
                              c(0,1), c("no", "yes"))),
            uterine.irr = as.factor(mapvalues(uterine.irr, 
                              c(0,1), c("no", "yes")))
            )
```

### ANOVA with birthwt data

**(a)** Create a new factor that categorizes the number of physician visits into three levels: 0, 1, 2, 3 or more.  

```{r}
phys.visit.binned <- birthwt$physician.visits
phys.visit.binned[phys.visit.binned >= 3] <- "3.or.more"
birthwt <- transform(birthwt, phys.visit.binned = as.factor(phys.visit.binned))

birthwt$phys.visit.binned
plot(birthwt$phys.visit.binned)

View(birthwt)
```

**Hint**: One way of doing this is with mapvalues, by mapping all instances of 3, 4,... etc, to "3 or more".

**(b)** Run an ANOVA to determine whether the average birth weight varies across number of physician visits.
```{r}
aov.birthwt3 <- aov(birthwt.grams ~ phys.visit.binned, data = birthwt, conf.level = 0.90)
aov.birthwt3
```
```{r}
aov.birthwt <- aov(birthwt.grams ~ phys.visit.binned, data = birthwt)
summary(aov.birthwt)
```

```{r}
aov.birthwt3 <- aov(birthwt.grams ~ phys.visit.binned, data = birthwt, conf.level = 0.90)
aov.birthwt3
```


```{r}
aov.birthwt_race <- aov(birthwt.grams ~ race, data = birthwt)
summary(aov.birthwt_race)
```


```{r}
aov.birthwt_race <- aov(birthwt.grams ~ race, data = birthwt)
summary(aov.birthwt_race)
plot(aov.birthwt_race)
#aov.birthwt
```

The p-value is greater than 0.05, so the variation in birthweight across number of physician visits is not statistically significant.  

```{r}
plot(aov.birthwt2)
```

```{r}
birthwt.lm <- lm(birthwt.grams ~ mother.age + mother.smokes, data = birthwt)
```


```{r}
birthwt.lm <- lm(birthwt.grams ~ mother.age + mother.smokes, data = birthwt)
summary(birthwt.lm)
```









```{r}
posthoc <- TukeyHSD(x=aov.birthwt, 'phys.visit.binned', conf.level=0.90)
posthoc
plot(posthoc)
```