---
title: "Regession"
author: "Manirath Wongsim"
date: ''
output: pdf_document
---

We'll begin by loading all the packages we might need.

```{r}
library(plyr)
```


```{r}
library(MASS)
library(plyr)
library(reshape)  # You may need to install this one first!
library(ggplot2)
```

### Is the data normal?

**(a)** Construct histograms of `MPG.highway`, one plot for each `Origin` category.

```{r}
qplot(x = MPG.highway, data = Cars93, facets = ~Origin, geom = "histogram", fill = Origin, binwidth = 2)
```

**(b)** Does the data look to be normally distributed?

The histograms don't really look normally distributed, so we might be better off using the non-parametric test.

**(c)** Construct qqplots of `MPG.highway`, one plot for each `Origin` category.  Overlay a line on each plot using with `qqline()` function.

```{r, fig.height = 4}
par(mfrow = c(1,2))
# USA cars
with(Cars93, qqnorm(MPG.highway[Origin == "USA"]))
with(Cars93, qqline(MPG.highway, col = "blue"))
# Foreign cars
with(Cars93, qqnorm(MPG.highway[Origin == "non-USA"]))
with(Cars93, qqline(MPG.highway, col = "blue"))
```

**(d)** Does the data look to be normally distributed?

The non-USA MPG.highway data looks very far from normally distributed.


### Testing means between two groups

**(a)** Using the Cars93 data and the `t.test()` function, run a t-test to see if average `MPG.highway` is different between US and non-US vehicles.

```{r}
View(Cars93)
library(plyr)
```

```{r}
View(survey_data)
```


```{r}
Program.t.test <- t.test(TVhours ~ OperatingSystem, data = survey)
Program.t.test
```

```{r}
Program.lm <- lm(TVhours ~ OperatingSystem + Program + PriorExp, data = survey_data)
summary(Program.lm)
```

Try doing this both using the formula style input and the `x`, `y` style input.

```{r}
View(Cars93)
```

```{r}
qplot(data = Cars93, x = EngineSize, y = Price, colour = Origin) + stat_smooth(method = "lm", fullrange = TRUE)
```

```{r}
library(plot3D)
R <- 5
r <- 2
x <- seq(0, 2*pi,length.out=20)
y <- seq(0, pi,length.out=80)
M <- mesh(x, y)car93.lm <- lm(MPG.highway ~ Origin + EngineSize + Cylinders, data = Cars93)
alpha <- M$x
beta <- M$y
surf3D(x = (R + r*cos(alpha)) * cos(beta),
       y = (R + r*cos(alpha)) * sin(beta),
       z = r * sin(alpha),
       colkey=FALSE,
       bty="b2",
       main="Big data with Hadoop #6")
```

```{r}
par(mfrow = c(2, 2), mar = c(0, 0, 0, 0))
# Shape 1
M  <- mesh(seq(0,  6*pi, length.out = 80),
seq(pi/3, pi, length.out = 80))
u  <- M$x ; v <- M$y
x <- u/2 * sin(v) * cos(u)
y <- u/2 * sin(v) * sin(u)
z <- u/2 * cos(v)
surf3D(x, y, z, colvar = z, colkey = FALSE, box = FALSE)
# Shape 2: add border
M  <- mesh(seq(0, 2*pi, length.out = 80),
seq(0, 2*pi, length.out = 80))
u  <- M$x ; v  <- M$y
x  <- sin(u)
y  <- sin(v)
z  <- sin(u + v)
surf3D(x, y, z, colvar = z, border = "black", colkey = FALSE)
# shape 3: uses same mesh, white facets
x <- (3 + cos(v/2)*sin(u) - sin(v/2)*sin(2*u))*cos(v)
y <- (3 + cos(v/2)*sin(u) - sin(v/2)*sin(2*u))*sin(v)
```


```{r}
car93.lm <- lm(MPG.highway ~ Origin +Horsepower +EngineSize + Cylinders, data = Cars93)
summary(car93.lm)
plot(car93.lm)
```

```{r}
mpg.test <- t.test(MPG.highway ~ Origin, data = Cars93)
mpg.test
```


```{r}
mpg.t.test <- t.test(MPG.highway ~ Origin, data = Cars93, conf.level = 0.90)
mpg.t.test

```


```{r}
car93.lm$coefficients
```

```{r}
# Formula version
mpg.t.test <- t.test(MPG.highway ~ Origin, data = Cars93)
mpg.t.test


```at confident level = 90%
## 
```{r}
mpg.t.test <- t.test(MPG.highway ~ Origin, data = Cars93, conf.level = 0.90)
mpg.t.test
```
```{r}
with(Cars93, t.test(x = MPG.highway[Origin == "USA"], y = MPG.highway[Origin == "non-USA"]))
```

```{r}
with(Cars93, t.test(x = MPG.highway[Origin == "USA"], y = MPG.highway[Origin == "non-USA"]))
```


```{r}
mpg.t.test$conf.int
```
```{r}
mpg.wilcox.test <- wilcox.test(MPG.highway ~ Origin, data = Cars93)
mpg.wilcox.test
```


# x, y version
with(Cars93, t.test(x = MPG.highway[Origin == "USA"], y = MPG.highway[Origin == "non-USA"]))
```

**(b)** What is the confidence interval for the difference?

```{r}
mpg.t.test$conf.int
```

**(c)** Repeat part (a) using the `wilcox.test()` function.

```{r}
mpg.wilcox.test <- wilcox.test(MPG.highway ~ Origin, data = Cars93)
mpg.wilcox.test
```


